National Longitudinal Evidence for Growth in Subjective Well-being from Spiritual Beliefs

### 2021-02-26 Manuscript_Bayesian.Rmd
fixes

abstract: | 
  Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage nine waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 20979, New Zealand, 2010-2018). Contrary to previous research, we find that belief in a spirit or Life Force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or Life Force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being. 
  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "5161 words in text body, 1191 words in reference section"
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]

---
```{r libraries}
library("tidyverse")
library("papaja")
library("here")

```
```{r global_options, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(#fig.path="Figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                     # fig.width= 10,
                     # fig.height=8,
                      tidy = "styler")
```

@@

The NZAVS is an annual, longitudinal national probability sample of registered New Zealand voters, which was started in 2009. The Time 10 wave of the NZAVS contained responses from 47,951 participants (17,981 retained from one or more previous waves and 29,970 new additions from booster sampling and/or unmatched participants or unsolicited opt-ins). Participants who provided an email address were also emailed and invited to complete an online version if they preferred. We offered a prize draw for participation, non-respondents were emailed and phoned multiple times, and all participants were mailed a Season’s Greetings card from the NZAVS research team and informed that they had been automatically entered into a bonus seasonal grocery voucher prize draw. We also mailed our yearly pamphlet summarizing key research findings published during the current wave of the study.

@@

The Time 10 (2018) wave of the NZAVS included 47,951 respondents. We analyzed data from participants who responded to our survey at least three times between Time 2 (2010) and Time 10 (2018), resulting in a sample of $N$ = 21,705 ($N$ = 20,979 with complete responses). Descriptive summaries for  demographic and theoretical indicators are given in Table 1.

Table 1 about here 
\begin{table}[h]
\begin{center}
\scalebox{0.45}{
\begin{tabular}[t]{llllllllllllllllllll}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{2010 (N=3820)} & \multicolumn{2}{c}{2011 (N=6153)} & \multicolumn{2}{c}{2012 (N=10385)} & \multicolumn{2}{c}{2013 (N=15836)} & \multicolumn{2}{c}{2014 (N=15026)} & \multicolumn{2}{c}{2015 (N=13665)} & \multicolumn{2}{c}{2016 (N=18071)} & \multicolumn{2}{c}{2017 (N=15779)} & \multicolumn{2}{c}{2018 (N=16501)} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10} \cmidrule(l{3pt}r{3pt}){11-12} \cmidrule(l{3pt}r{3pt}){13-14} \cmidrule(l{3pt}r{3pt}){15-16} \cmidrule(l{3pt}r{3pt}){17-18} \cmidrule(l{3pt}r{3pt}){19-20}
  &    & Mean & Std. Dev. & Mean  & Std. Dev.  & Mean   & Std. Dev.   & Mean    & Std. Dev.    & Mean     & Std. Dev.     & Mean      & Std. Dev.      & Mean       & Std. Dev.       & Mean        & Std. Dev.        & Mean         & Std. Dev.        \\
\midrule
Age &  & 51.0 & 14.7 & 50.9 & 15.5 & 50.2 & 14.8 & 48.6 & 14.0 & 49.7 & 14.0 & 51.0 & 13.9 & 50.9 & 13.6 & 51.9 & 13.6 & 52.7 & 13.5\\
NZdep &  & 4.8 & 2.8 & 4.6 & 2.7 & 4.9 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.7 & 2.8 & 4.6 & 2.7 & 4.6 & 2.7 & 4.6 & 2.7\\
Education &  &  &  &  &  & 5.8 & 2.8 & 6.0 & 2.8 & 6.1 & 2.8 & 6.2 & 2.8 & 6.4 & 2.7 & 6.5 & 2.7 & 6.5 & 2.7\\
Political\_Conservativism &  & 3.9 & 1.3 & 3.7 & 1.4 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.3 & 3.6 & 1.4 & 3.6 & 1.4 & 3.6 & 1.4\\
LIFESAT &  & 5.2 & 1.2 & 5.1 & 1.2 & 5.1 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.2 & 1.2 & 5.3 & 1.2 & 5.3 & 1.2\\
PWI &  & 6.8 & 1.7 & 6.8 & 1.7 & 6.8 & 1.7 & 7.0 & 1.7 & 7.0 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.7 & 7.1 & 1.6\\
\midrule
 &  & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \% & N & \%\\
Employed & 0 & 793 & 0.7 & 1276 & 1.1 & 2666 & 2.3 & 3531 & 3.1 & 3262 & 2.8 & 2972 & 2.6 & 4035 & 3.5 & 3655 & 3.2 & 3746 & 3.3\\
 & 1 & 2977 & 2.6 & 4795 & 4.2 & 7566 & 6.6 & 11991 & 10.4 & 11626 & 10.1 & 10378 & 9.0 & 14016 & 12.2 & 12070 & 10.5 & 12250 & 10.6\\
Ethnic\_Categories & Euro & 2947 & 2.6 & 4375 & 3.8 & 7881 & 6.8 & 12270 & 10.6 & 12051 & 10.5 & 10959 & 9.5 & 14641 & 12.7 & 12906 & 11.2 & 13475 & 11.7\\
 & Maori & 556 & 0.5 & 638 & 0.6 & 1489 & 1.3 & 1897 & 1.6 & 1830 & 1.6 & 1608 & 1.4 & 1964 & 1.7 & 1826 & 1.6 & 1872 & 1.6\\
 & Pacific & 110 & 0.1 & 126 & 0.1 & 343 & 0.3 & 408 & 0.4 & 400 & 0.3 & 336 & 0.3 & 385 & 0.3 & 286 & 0.2 & 296 & 0.3\\
 & Asian & 124 & 0.1 & 190 & 0.2 & 409 & 0.4 & 579 & 0.5 & 572 & 0.5 & 473 & 0.4 & 664 & 0.6 & 574 & 0.5 & 579 & 0.5\\
Male & 0 & 2390 & 2.1 & 3862 & 3.4 & 6542 & 5.7 & 10033 & 8.7 & 9500 & 8.2 & 8548 & 7.4 & 11377 & 9.9 & 9949 & 8.6 & 10388 & 9.0\\
 & 1 & 1430 & 1.2 & 2291 & 2.0 & 3843 & 3.3 & 5803 & 5.0 & 5478 & 4.8 & 5072 & 4.4 & 6635 & 5.8 & 5776 & 5.0 & 6077 & 5.3\\
Has\_Partner & 0 & 1054 & 0.9 & 1806 & 1.6 & 2988 & 2.6 & 4237 & 3.7 & 3664 & 3.2 & 3407 & 3.0 & 4274 & 3.7 & 3775 & 3.3 & 3931 & 3.4\\
 & 1 & 2764 & 2.4 & 4325 & 3.8 & 7331 & 6.4 & 11493 & 10.0 & 11130 & 9.7 & 10176 & 8.8 & 13613 & 11.8 & 11964 & 10.4 & 12078 & 10.5\\
Urban & Not\_Urban & 1531 & 1.3 & 2010 & 1.7 & 3512 & 3.0 & 5116 & 4.4 & 4894 & 4.2 & 4541 & 3.9 & 6223 & 5.4 & 2873 & 2.5 & 2938 & 2.5\\
 & Urban & 2271 & 2.0 & 3695 & 3.2 & 6709 & 5.8 & 10602 & 9.2 & 9937 & 8.6 & 8815 & 7.6 & 11622 & 10.1 & 12687 & 11.0 & 13291 & 11.5\\
Beliefs & \_Disbeliefs\_ & 625 & 0.5 & 1312 & 1.1 & 2235 & 1.9 & 3639 & 3.2 & 0 & 0.0 & 3278 & 2.8 & 4420 & 3.8 & 4135 & 3.6 & 4514 & 3.9\\
 & \_SpiritExcludesGod\_ & 778 & 0.7 & 1508 & 1.3 & 2535 & 2.2 & 4206 & 3.6 & 0 & 0.0 & 3542 & 3.1 & 4808 & 4.2 & 4148 & 3.6 & 4309 & 3.7\\
 & GodAndSpirit & 1751 & 1.5 & 2560 & 2.2 & 4181 & 3.6 & 6201 & 5.4 & 0 & 0.0 & 5780 & 5.0 & 7002 & 6.1 & 5771 & 5.0 & 6782 & 5.9\\
 & GodExcludesSpirit & 228 & 0.2 & 274 & 0.2 & 593 & 0.5 & 1137 & 1.0 & 0 & 0.0 & 523 & 0.5 & 1094 & 0.9 & 950 & 0.8 & 666 & 0.6\\
\bottomrule
\end{tabular}
}
\caption{Sample statistics by year}
\end{center}
\end{table}

## Theoretical indicators 
### Life Satisfaction
Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy]. Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach’s alpha for this 2-item scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items.

@@

### Age 
Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2.
Gender. Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0."

### Education
Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree). Education was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

### Deprivation
We measured the socioeconomic status of participants’ immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex]. The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80–90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

### Employed
Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

### Partner
Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

### Ethnicity
Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

### Urban
People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.

### Political Orientation
To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative). Higher values indicate more conservative political beliefs. Political Orientation was standardized and centered at its mean. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

### Personal well-being 
Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur]. Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items.

### Belief in a Spirit/Life Force and/or a God  
We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as 1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: those who believe in a both God and a Spirit or Life Force.

## Statistical Analyses
Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least three times between 2010 (Time 2) and 2018 (Time 10), resulting in a sample of  21,705 (*N* = 20,979 with complete responses).  To model growth in subjective well-being we use generalized linear mixed models. In our models, we account for the non-linear effects of time within participants by including the interaction of years (a continuous variable) with Beliefs (a four-level categorical indicator). The baseline for this indicator is disbelief in a Spirit/Life Force and disbelief in a God. To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for years. This allowed us to model correllation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years. We estimated all statitiscal models using both REML estimation [@R-lme4] and Bayesian estimation [@BRMSpackage] with weakly informative priors. We include the full reports in the Supplement. The REML estimates and the Bayesian estimates converged to within three decimal places.  We include model equations and priors at the end of this article. Here we report the results of the Bayesian estimator.

@@

To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf). Next we modelled potential relationships between the coefficients within the model using the daggity package in R, and plotting results using the daggity and ggdag packages in R.

This method reveals two adjustments sets equired to avoid potential confounding: \{age, employment, male, political conservativism, and time\}, and \{age, ethnicty, male, and time \}.Here, we report the second set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g. retirement).  Results were insensitive to the choice set. Indeed, the inference proved robust to the simple model with only the interaction of years and beliefs, and robust to the fully saturated model in which all covariates were included (see the online supplement for details). In the fully saturated model, it appeared that those disbelieving in a Spirit or Life Force, exhibited no growth in Persona Well-Being and Life-Satisifaction over the nine years. Although this result offers stronger stronger evidence for a Spirit/Life Force well-being mechanism, it is likely the effect of collider confounding and masking. We discuss this point in more detail in the supplement, and mention it here to caution researchers from "everything-but-the-kitchen-sink" approaches to regression modelling.

## Missing Data 
Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences. Additionally the results were robust to filtering of participants of N > 3 to N = 8 waves. 
We include the multiple-imputation results in a supplement. Here, there were no differences in estimation relevant to inference, however, to improve inference, we advocate wider use of multiple-imputation or Bayesian approaches for handling missingness.

## System and packages
The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6
We are greatful to the contributors and mantainers of the following packages:
R [Version 4.0.3; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 1.0.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.25; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.9.0.1; @R-parameters], *patchwork* [Version 1.1.0; @R-patchwork], *prettycode* [@R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.4.0; @R-readr], *report* [Version 0.2.0; @R-report], *see* [Version 0.6.0.1; @R-see], *sjPlot* [Version 2.8.6; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2.1; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]

## Code
All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing](https://github.com/jacopastorius/nzavs_spiritbeliefWellbeing). An html document decribing the same analysis is located at: [https://osf.io/cu2gr/](https://osf.io/cu2gr/).

@@

We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict personal well-being (PWI) with years (0-9), beliefs (four categories), with Id as random intercept and years as a random slope effect (see model equations, below).

>>>>

abstract: | 
  Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological spirituality scales, poorly defined concepts of spirituality, and heavy reliance on cross-sectional samples cast doubts. Here, we leverage ten waves of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ = 3,257, New Zealand, 2010-2020). Contrary to previous research, we find that belief in a Spirit or Life Force predicts lower personal well-being and life-satisfaction. However, in support of previous speculation, beliefs in a spirit or Life Force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding are robust after to known demographic influences, and intriguingly, hold among those who believe in a God but disbelieve in a Spirit or Life Force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates further causal investigations of the mechanisms by which spiritual beliefs lead to growth in subjective well-being. 
  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "4760 words in text body, 1191 words in reference section"
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable","amsmath","amsfonts","amssymb"]
---

```{r libraries}
library("tidyverse")
library("papaja")
library("here")

```

```{r global_options, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(#fig.path="Figs/", 
                      message=FALSE, 
                      warning=FALSE,
                      collapse =TRUE,
                      echo=TRUE, 
                      #results="hide", 
                     # fig.width= 10,
                     # fig.height=8,
                      tidy = "styler")
```

@@

The NZAVS is an annual, longitudinal national probability sample of registered New Zealand voters, which was started in 2009. The NZAVS samples size by waves is as follows: Wave 1 (2009/10) = 6518; Wave 2 (2010/11) = 4441; Wave 3 (2011/12) = 6884; Wave 4 (2012/13) =  12179; Wave 5 (2013/14) =  18,261; Wave 6 (2014/15) = 15,820; Wave 7 (2015/16) = 13,942; Wave 8 (2016/17) = 21,936; Wave 9 (2017/18) =  17,072; Wave 10 (2018/19) = 47951; Wave 11(2019/20) = 36,524.

The first NZAVS measures of belief in God and belief in a spirit or life force appear in Wave 2 (2010/11) NZAVS contained responses from 4,441 participants. The Wave 2 (2010/11) NZAVS retained 4,425 from the initial Time 1 (2009/10) NZAVS sample of 6,518 participants, and included an additional 16 respondents who could not be matched to the Time 1 participant database (a retention rate of 67.9% over one year). Participants in the initial Time 1 (2009) sample were randomly selected from the New Zealand electoral roll (a national registry of registered voters). The response rate in the initial Time 1 sample, adjusting for the accuracy of the electoral roll and including anonymous responses was 16.6%. Participants were posted a copy of the questionnaire, with a second postal follow-up two months later. Participants who provided an email address were also emailed and invited to complete an online questionnaire if they preferred.  Full details for the NZAVS sampling procedure can be found in the supplement to this study can be found on our OSF link: <https://osf.io/cu2gr/>

@@

We analyzed data from participants who responded to our survey in at least 9 of the 10 waves between Time 2 (2010/2011) and Time 11 (2019/2020), resulting in a sample of $N$ = 3,257 individuals. Descriptive summaries for demographic and theoretical indicators are given in Table 1.

Table 1 about here.

```{=tex}
\begin{table}
\begin{center}
\scalebox{0.8}{
\begin{tabular}{l c c}
\toprule
 & PWI & Life Sat \\
\midrule
Intercept                            & $\mathbf{7.24}$   & $\mathbf{5.36}$   \\
                                     & $ [ 7.18;  7.30]$ & $ [ 5.31;  5.40]$ \\
BeliefsGodAndSpirit                  & $\mathbf{-0.08}$  & $-0.01$           \\
                                     & $ [-0.13; -0.03]$ & $ [-0.05;  0.02]$ \\
BeliefsGodExcludesSpirit             & $-0.04$           & $0.01$            \\
                                     & $ [-0.10;  0.03]$ & $ [-0.04;  0.06]$ \\
BeliefsSpiritExcludesGod             & $\mathbf{-0.06}$  & $-0.02$           \\
                                     & $ [-0.10; -0.01]$ & $ [-0.05;  0.01]$ \\
Age\_within                          & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
                                     & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
Age\_betweenC                        & $\mathbf{0.01}$   & $\mathbf{0.01}$   \\
                                     & $ [ 0.01;  0.01]$ & $ [ 0.01;  0.01]$ \\
Ethnic\_CategoriesMaori              & $-0.08$           & $0.04$            \\
                                     & $ [-0.16;  0.00]$ & $ [-0.02;  0.10]$ \\
Ethnic\_CategoriesPacific            & $\mathbf{-0.39}$  & $\mathbf{-0.29}$  \\
                                     & $ [-0.54; -0.24]$ & $ [-0.39; -0.18]$ \\
Ethnic\_CategoriesAsian              & $-0.05$           & $-0.04$           \\
                                     & $ [-0.20;  0.13]$ & $ [-0.15;  0.09]$ \\
Male1                                & $-0.05$           & $\mathbf{-0.11}$  \\
                                     & $ [-0.13;  0.04]$ & $ [-0.16; -0.05]$ \\
BeliefsGodAndSpirit:Age\_within      & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
                                     & $ [ 0.01;  0.03]$ & $ [ 0.01;  0.02]$ \\
BeliefsGodExcludesSpirit:Age\_within & $0.01$            & $0.00$            \\
                                     & $ [-0.01;  0.03]$ & $ [-0.01;  0.02]$ \\
BeliefsSpiritExcludesGod:Age\_within & $\mathbf{0.02}$   & $\mathbf{0.01}$   \\
                                     & $ [ 0.01;  0.03]$ & $ [ 0.00;  0.02]$ \\
\midrule
SD: Idsd(Intercept)                  & $1.37$            & $0.96$            \\
SD: Idsd(Age\_within)                & $0.10$            & $0.08$            \\
SD: Idcor(Intercept,Age\_within)     & $-0.05$           & $-0.01$           \\
\bottomrule
\end{tabular}
}
\caption{}
\label{tab:REGRESS_LS}
\end{center}
\end{table}
```
## Theoretical indicators

### Life Satisfaction

Life satisfaction is a measure of emotional well-being that was assessed using a 2-item version of the Satisfaction With Life Scale, which has previously been shown to correlate with aspects of religiosity [@Diener1985-xy]. Participants rated their agreement with the statements (a) "I am satisfied with my life"; and (b) "In most ways my life is close to ideal." The items were rated on 7-point response options ranging from 1 = strongly disagree to 7 = strongly agree. The mean Cronbach's alpha for this 2-item scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher life satisfaction. Overall scores were means of the 2-items.

@@

### Age

Age was put into units of 10 years and centered at its mean. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 2. Gender. Gender was assessed by asking participants if they were "Male" was coded as "1" and "Female" was coded as "0."

### Education

Education level was measured using an 11-point rating developed by the New Zealand Qualification Authority known as the New Zealand Qualification Framework (NZQF; 0 = no qualification, 10 = doctoral degree). Education was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

### Deprivation

We measured the socioeconomic status of participants' immediate (small area) neighborhood using the 2013 New Zealand Deprivation Index, which uses aggregate census information about the residents of small neighborhood-type units to assign a decile-rank index from 1 (most affluent) to 10 (most impoverished) [@Atkinson2014-ex]. The index is based on a Principal Components Analysis of the following nine variables (in weighted order): the proportion of adults who received a means-tested benefit, household income, proportion not owning own home, proportion single-parent families, proportion unemployed, proportion lacking qualifications, proportion household crowding, proportion no telephone access, and proportion no car access. Thus, the index reflects the average level of deprivation for small neighborhood-type units (or small community areas of about 80--90 people each) across the entire country. Our sample had a mean deprivation index of 4.80 (SD = 2.79). Deprivation was centered at its mean and standardized. For detailed information pertaining to yearly means, standard deviations, and missingness, see Table 1.

### Employed

Employment status was assessed by asking participants if they were currently working, "yes" was coded as "1" and "no" was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

### Partner

Participants were asked if they were in a relationship, "yes" was coded as "1" and no was coded as "0." For detailed information pertaining to yearly responses and missingness, see Table 1.

### Ethnicity

Ethnicity was assessed using four basic categories: (1) New Zealand European/Pakeha, (2) Maori, (3) Pacific Islander, and (4) Asian. For detailed information pertaining to yearly responses and missingness, see Table 1.

### Urban

People were coded as either residing in an urban "1" or rural "0" area based on New Zealand census data. For detailed information pertaining to yearly responses and missingness, see Table 1.

### Political Orientation

To assess political orientation, we asked people to rate their political orientation using seven-point response options (1 = Liberal; 7 = Conservative). Higher values indicate more conservative political beliefs. Political Orientation was standardized and centered at its mean. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

### Personal well-being

Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur]. Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach's alpha for this scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items.

### Belief in a Spirit/Life Force and/or a God

We assess belief in a spirit or a God by asking two non-affective questions: "Do you believe in a spirit or Life Force" and "Do you believe in a god?" Responses were coded as 1) **God and Spirit Disbelievers**: those who neither believe in a God nor believe in a Spirit or Life Force; (2) **Spirit Believers/God Disbelievers**: those who disbelieve in a God and believe in a Spirit or Life Force; (3) **God Believers/Spirit Disbelievers**: those who believe in a God but do not believe in a Spirit or Life Force; (4) **God and Spirit Believers**: those who believe in a both God and a Spirit or Life Force.

## Statistical Analyses

Statistical analysis was performed using R version 4.0.2 R version 3.6.1 (2020-06-22) 2019-07-05). We analyze data from participants who responded to the NZAVS survey at least nine times between 2010/11 (Time 2) and 2019/20 (Time 11), resulting in a sample of 21,705 ($N$ = 3,257 with complete responses). To model growth in subjective well-being we use generalized linear mixed models. To avoid heterogeneity bias, we "demeaned" the variable "age" separating this variable into the within and between person components. Heterogeneity bias arises when an indicators varies within and between groups, which will case fixed effects estimates to correlate with group-level effects [@bell2015explaining], which is a common feature of panel data (in which a participant is a level-2 indicator)[@R-parameters]. We modelled the interaction of within participant aging and belief by including the interaction of aging within (a continuous variable, centered) and belief-state, a four-level categorical indicator. The baseline for this indicator is disbelief in a Spirit/Life Force and disbelief in a God -- or "disbeliever." To handle the dependencies introduced from the repeated measures, we include individual ID as an effect modeled as random, and we estimated a random effect for the slope for aging (within) This approach allowed us to model correlation between individual-level intercept (starting high or low in Personal Well-Being/Life Satisfaction) and stability/growth/decline in the slope effect of years. We estimated all statistical models using both REML estimation [@R-lme4] and Bayesian estimation [@BRMSpackage] with weakly informative priors, reported in the on-line supplement. The REML estimates, which we report in the supplement, converged to within three decimal places. The model equations and priors are in the on-line supplement.

@@

To address confounding we first obtained the New Zealand Attitudes and Values Study standard demographic indicators recommended to address confounding: [NZAVS modelling standards](https://cdn.auckland.ac.nz/assets/psych/about/our-research/nzavs/NZAVSTechnicalDocuments/NZAVS-Technical-Documents-e11-Default-Statistical-Standards.pdf). Next we modelled potential relationships between the coefficients within the model using the `ggdag` package in R.

This method reveals two adjustments sets required to avoid potential confounding: \{age_within, age_between, employment, male, political conservativism, and time\}, and \{age_within, age_between, ethnicty, male, and time\}. Here, we used the second adjustment set because employment is likely to contain more measurement error: some are not employed because they cannot find employment, others because they are not seeking employment (e.g. retirement). Results were insensitive to the choice set.

## Missing Data

Because missing values can result in biased estimates [@Blackwell2017-oq], we multiply imputed 10 datasets using the AmeliaII package in R[@Honaker2011-yu].We conducted a parallel analysis averaging over the multiply imputed datasets, without any substantive difference to inferences.

## System and packages

The analysis was performed using R version 4.0.2 (2020-06-22). The Platform was x86_64-apple-darwin17.0 (64-bit) Running under: macOS Catalina 10.15.6 We are greatful to the contributors and mantainers of the following packages: R [Version 4.0.3; @R-base] and the R-packages *Amelia* [Version 1.7.6; @R-Amelia], *citr* [Version 0.3.2; @R-citr], *dplyr* [Version 1.0.2; @R-dplyr], *equatiomatic* [Version 0.1.0.9000; @R-equatiomatic], *forcats* [Version 0.5.0; @R-forcats], *ggeffects* [Version 1.0.0; @R-ggeffects], *gghighlight* [Version 0.3.0; @R-gghighlight], *ggplot2* [Version 3.3.2; @R-ggplot2], *ggsci* [Version 2.9; @R-ggsci], *lme4* [Version 1.1.25; @R-lme4], *Matrix* [Version 1.2.18; @R-Matrix], *papaja* [Version 0.1.0.9997; @R-papaja], *parameters* [Version 0.9.0.1; @R-parameters], *patchwork* [Version 1.1.0; @R-patchwork], *prettycode* [@R-prettycode], *purrr* [Version 0.3.4; @R-purrr], *Rcpp* [Version 1.0.5; @R-Rcpp_a; @R-Rcpp_b], *readr* [Version 1.4.0; @R-readr], *report* [Version 0.2.0; @R-report], *see* [Version 0.6.0.1; @R-see], *sjPlot* [Version 2.8.6; @R-sjPlot], *stringr* [Version 1.4.0; @R-stringr], *styler* [Version 1.3.2; @R-styler], *table1* [Version 1.2.1; @R-table1], *texreg* [Version 1.37.5; @R-texreg], *tibble* [Version 3.0.4; @R-tibble], *tidyr* [Version 1.1.2; @R-tidyr], and *tidyverse* [Version 1.3.0; @R-tidyverse]

## Code

All code for this analysis and a full report of the missing data imputation can be found at [https://github.com/go-bayes/nzavs_spiritbeliefWellbeing](https://github.com/go-bayes/nzavs_spiritbeliefWellbeing){.uri}. An html document decribing the same analysis is located at: <https://osf.io/cu2gr/>.

@@

We fitted a linear mixed effects model using the brms package [@BRMSpackage] in R [@R-base] to predict Life Satisfaction (LS) with age_within centered (range = -5.79, 6.83), beliefs (four categories), with Id as random intercept and years as a random slope effect (see below: model equations).
